---
title: "Practice Exam"
author: "Colin Yip, Student No. 3953629"
date: "2023-12-14"

header-includes:
  \usepackage{geometry}
  \usepackage{amsmath}
  \geometry{top=0.25in, bottom=0.25in, left=0.25in, right=0.25in}

output:
  pdf_document:
    fig_crop:
      no
---

```{r global_set, echo = F}
library(knitr)
options(width = 80)
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(tidy_opts = list(width.cutoff = 80))
```

# Exercise 1
## Q1
```{r}
car_df <- read.csv("cartype.csv")

head(car_df)
```
The data is in a long format, as it is year over year per row.


## Q2
```{r}
library(dplyr)
target_countries <- c("France", "Germany", "Italy", "Spain", "United Kingdom")
filtered_car_df <- dplyr::filter(car_df, Entity %in% target_countries)

head(filtered_car_df)
```

## Q3
```{r}
filtered_car_df$total <- filtered_car_df$battery_electric_number +
  filtered_car_df$petrol_number +
  filtered_car_df$diesel_gas_number +
  filtered_car_df$hybrid_number
total_per_year <- filtered_car_df %>%
  group_by(Year) %>%
  summarize(all_numbers = sum(total))
total_per_type_per_year <- filtered_car_df %>%
  group_by(Year) %>%
  summarize(
    batt = sum(battery_electric_number),
    petrol = sum(petrol_number),
    diesel = sum(diesel_gas_number),
    hybrid = sum(hybrid_number)
  )
batt_prop <- cbind(
  total_per_year$Year,
  total_per_type_per_year$batt / total_per_year$all_numbers
)
petrol_prop <- cbind(
  total_per_year$Year,
  total_per_type_per_year$petrol / total_per_year$all_numbers
)
diesel_prop <- cbind(
  total_per_year$Year,
  total_per_type_per_year$diesel / total_per_year$all_numbers
)
hybrid_prop <- cbind(
  total_per_year$Year,
  total_per_type_per_year$hybrid / total_per_year$all_numbers
)
```

## Q4
```{r}
filtered_cars_2008 <- filter(filtered_car_df, Year == 2008)
max_diesel_idx <- which.max(filtered_cars_2008$diesel_gas_number)
max_diesel_2008 <- filtered_cars_2008[max_diesel_idx, "Entity"]

filtered_cars_2018 <- filter(filtered_car_df, Year == 2018)
max_diesel_idx <- which.max(filtered_cars_2018$diesel_gas_number)
max_diesel_2018 <- filtered_cars_2018[max_diesel_idx, "Entity"]
```
Maximum diesel car country in 2008 was `r max_diesel_2008` and maximum diesel car country in 2018 was `r max_diesel_2018`.

## Q5
```{r}
filtered_cars_2005 <- filter(filtered_car_df, Year == 2005, Entity == "Germany")
engine_type_names <- c(
  "battery_electric_number", "petrol_number",
  "diesel_gas_number", "hybrid_number"
)
pie(as.numeric(filtered_cars_2005[, engine_type_names]),
  col = c("blue", "red", "purple", "orange"),
  main = "2005 Germany Cars by Engine Type",
  labels = engine_type_names
)

filtered_cars_2018 <- filter(filtered_car_df, Year == 2018, Entity == "Germany")
engine_type_names <- c(
  "battery_electric_number", "petrol_number",
  "diesel_gas_number", "hybrid_number"
)
pie(as.numeric(filtered_cars_2018[, engine_type_names]),
  col = c("blue", "red", "purple", "orange"),
  main = "2018 Germany Cars by Engine Type",
  labels = engine_type_names
)
```
The proportion of diesel and petrol cars has reduced, with the proportion of hybrid and electric cars increasing.

### Q6
```{r}
# Group car numbers by country
hybrid_car_by_country <- filtered_car_df %>%
  group_by(Year, Entity) %>%
  summarize(
    hybrid_number = hybrid_number,
    total_cars = sum(
      battery_electric_number,
      petrol_number,
      diesel_gas_number,
      hybrid_number
    )
  )
# Get proportion by country, year
hybrid_car_by_country$hybrid_prop <- hybrid_car_by_country$hybrid_number /
  hybrid_car_by_country$total_cars

library(ptmixed)
make.spaghetti(Year, hybrid_prop,
  group = Entity, id = Entity,
  data = data.frame(hybrid_car_by_country)
)
```
Spain and UK have fastest diffusion of hybrids.

## Q7
```{r}
library(reshape2)
france_data <- filter(filtered_car_df, Entity == "France")
france_car_types <- france_data %>%
  group_by(Entity, Year) %>%
  summarise(
    battery_electric_number = battery_electric_number,
    petrol_number = petrol_number,
    diesel_gas_number = diesel_gas_number,
    hybrid_number = hybrid_number
  )
melted_france_car_types <- melt(france_car_types,
  id.vars = c("Entity", "Year")
)

make.spaghetti(Year, value,
  group = variable, id = variable,
  data = melted_france_car_types
)
```

# Exercise 2
## Q1
```{r}
k <- 7
lambda <- 2

ex2_1_func <- function(x) {
  return((k / lambda) * ((x / lambda)^(k - 1)) * (exp(-((x / lambda)^k))))
}

x_range <- seq(0, 10, 0.1)
curve(ex2_1_func, from = 0, to = 10)
```

## Q2
```{r}
k <- 3
lambda <- 4
x_range <- c(0.01, 30)
ex2_2_func <- function(x, k, lambda) {
  return((k / lambda) * ((x / lambda)^(k - 1)) * (exp(-((x / lambda)^k))))
}
maximized_x <- optimize(ex2_2_func, x_range, k, lambda, maximum = T)$maximum
```
The mode is `r maximized_x`.

## Q3
```{r}
k <- 2.3
lambda <- 5.7
x_range <- c(0.01, 30)
ex2_2_func <- function(x, k, lambda) {
  return((k / lambda) * ((x / lambda)^(k - 1)) * (exp(-((x / lambda)^k))))
}
maximized_x <- optimize(ex2_2_func, x_range, k, lambda, maximum = T)$maximum
```
The mode is `r maximized_x`.

## Q5
```{r}
mle_mode <- function(k, lambda) {
  return(lambda * ((k-1)/k) ^ (1/k))
}

mle_mode(3, 4)
mle_mode(2.3, 5.7)

```
Yes - advantage is that it is much more straightforward to compute numerically as it is compolicated to solve analytically. But analyticially is more precise.

# Exercise 3
## Q1
```{r}
getSummaries <- function(x) {
  n <- length(x)
  mean_vec <- mean(x)
  median_vec <- median(x)
  variance_vec <- var(x)

  values <- c(n, mean_vec, median_vec, variance_vec)

  statistic_names <- c(
    "Sample size",
    "Mean",
    "Median",
    "Variance"
  )

  output_df <- data.frame(cbind(statistic_names, values))
  names(output_df) <- c("Statistic", "Value")
  knitr::kable(output_df)
}

some_vec <- 1:10

getSummaries(some_vec)
```

## Q2
```{r}
getSummaries <- function(x, digits) {
  n <- length(x)
  mean_vec <- round(mean(x), digits)
  median_vec <- round(median(x), digits)
  variance_vec <- round(var(x), digits)

  values <- c(n, mean_vec, median_vec, variance_vec)

  statistic_names <- c(
    "Sample size",
    "Mean",
    "Median",
    "Variance"
  )

  output_df <- data.frame(cbind(statistic_names, values))
  names(output_df) <- c("Statistic", "Value")
  knitr::kable(output_df)
}

getSummaries(some_vec, 2)
```

## Q3
```{r}
getSummaries <- function(x, digits, ignoreNAs) {
  n <- length(x)
  if (NA %in% x & !ignoreNAs) {
    warning("There's NAs in x. Returning only n.")
    output_df <- cbind(c("Sample size"), c(n))
  } else if (NA %in% x & ignoreNAs) {
    mean_vec <- round(mean(x, na.rm = T), digits)
    median_vec <- round(median(x, na.rm = T), digits)
    variance_vec <- round(var(x, na.rm = T), digits)

    values <- c(n, mean_vec, median_vec, variance_vec)

    statistic_names <- c(
      "Sample size",
      "Mean",
      "Median",
      "Variance"
    )

    output_df <- data.frame(cbind(statistic_names, values))
  } else {
    mean_vec <- round(mean(x), digits)
    median_vec <- round(median(x), digits)
    variance_vec <- round(var(x), digits)

    values <- c(n, mean_vec, median_vec, variance_vec)

    statistic_names <- c(
      "Sample size",
      "Mean",
      "Median",
      "Variance"
    )

    output_df <- data.frame(cbind(statistic_names, values))
  }
  names(output_df) <- c("Statistic", "Value")
  knitr::kable(output_df)
}

vec_with_na <- c(1:5, NA, 7:10)
getSummaries(vec_with_na, 2, F)
```

## Q4
```{r}
getSummariesPerVec <- function(x, digits, ignoreNAs) {
  n <- length(x)
  if (NA %in% x & !ignoreNAs) {
    warning("There's NAs in x. Returning only n.")
    output_df <- cbind(c("Sample size"), c(n))
  } else if (NA %in% x & ignoreNAs) {
    mean_vec <- round(mean(x, na.rm = T), digits)
    median_vec <- round(median(x, na.rm = T), digits)
    variance_vec <- round(var(x, na.rm = T), digits)

    values <- c(n, mean_vec, median_vec, variance_vec)

    statistic_names <- c(
      "Sample size",
      "Mean",
      "Median",
      "Variance"
    )

    output_df <- data.frame(cbind(statistic_names, values))
  } else {
    mean_vec <- round(mean(x), digits)
    median_vec <- round(median(x), digits)
    variance_vec <- round(var(x), digits)

    values <- c(n, mean_vec, median_vec, variance_vec)

    statistic_names <- c(
      "Sample size",
      "Mean",
      "Median",
      "Variance"
    )

    output_df <- data.frame(cbind(statistic_names, values))
  }
  names(output_df) <- c("Statistic", "Value")
  knitr::kable(output_df)
}

getSummaries <- function(x, digits, ignoreNAs) {
  if (is.matrix(x)) {
    x_ncol <- ncol(x)
    x_ncol_range <- 1:x_ncol
    apply(x, 2, getSummariesPerVec, digits, ignoreNAs)
  } else if (is.vector(x)) {
    getSummariesPerVec(x, digits, ignoreNAs)
  }
}

some_matrix <- matrix(1:4, nrow = 2, ncol = 2)
getSummaries(some_matrix, 2, T)
```

## Q5
```{r}
getSummaries <- function(x, digits, ignoreNAs) {
  if (is.matrix(x)) {
    x_ncol <- ncol(x)
    x_ncol_range <- 1:x_ncol
    apply(x, 2, getSummariesPerVec, digits, ignoreNAs)
  } else if (is.vector(x) & is.numeric(x)) {
    getSummariesPerVec(x, digits, ignoreNAs)
  } else {
    stop("Neither matrix nor vector passed.")
  }
}
```

## Q6
```{r}
set.seed(3078)
x1 <- rhyper(300, 5, 20, 10)
x2 <- c(rep(cars$speed, 3), rep(NA, 50), rep(cars$dist, 2))
x3 <- cbind(x1, x2)

sapply(list(x1, x2, x3), getSummaries, digits = 2, ignoreNAs = T)
```

## Q7
```{r}
sapply(list(x1, x2, x3), getSummaries, digits = 4, ignoreNAs = F)