---
title:
  'Statisical Computing with R - Assignment 3'

subtitle:
  'Colin Yip, Student No. 3953629'
  
header-includes:
  \usepackage{geometry}
  \usepackage{amsmath}
  \geometry{top=0.5in, left=0.5in, right=0.5in, bottom=0.5in}
  
output:
  pdf_document:
    fig_crop: no
---
# Exercise 1
Formatting and output config.
```{r global_set, echo = TRUE}
library(knitr)
library(formatR)
library(palmerpenguins)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80))
```

### Q1
```{r}
# Read in penguins
penguins_data <- palmerpenguins::penguins
# Print penguins to inspect
print(penguins_data)
# Convert penguins to a dataframe
penguins_df <- as.data.frame(penguins_data)
```
`penguins` is a tibble.

### Q2
```{r}
# Count penguins by species and island
penguins_frequency_dist <- penguins_df %>%
  group_by(species, island) %>%
  count()
# Convert to DF
penguins_frequency_dist_df <- as.data.frame(penguins_frequency_dist)
# Print results
knitr::kable(penguins_frequency_dist_df,
  caption = "Distribution Frequency of Penguins by Species and Island"
)
```
There are `r length(unique(penguins_frequency_dist_df$species))` different species in this dataset.
The following are the species found on each island.

```{r}
# Get unique penguin categories by island
species_by_island_df <- penguins_df %>%
  group_by(island) %>%
  summarise(species = unique(species))
# Print results
knitr::kable(species_by_island_df,
  caption = "Penguin Species by Island"
)
```
### Q3
```{r}
# Filter for Gentoo penguins and get bill length values
gentoo_idx <- which(penguins_df$species == "Gentoo")
gentoo_bill_lengths <- penguins_df[gentoo_idx, ]$bill_length_mm

# Filter for Chinstrap penguins and get bill length values
chinstrap_idx <- which(penguins_df$species == "Chinstrap")
chinstrap_bill_lengths <- penguins_df[chinstrap_idx, ]$bill_length_mm

# Define significance value
alpha <- 0.05

# Do t test to evaluate if chinstrap average length
# is greater than gentoo average length
t_test_results <- t.test(
  x = chinstrap_bill_lengths,
  y = gentoo_bill_lengths,
  alternative = "less",
  var.equal = F
)
t_test_pvalue <- t_test_results$p.value
# Evaluate if p-value of t test is greater than signif. value
if (t_test_pvalue < alpha) {
  print("p-value of t-test is less than alpha, so we reject H0")
} else {
  print("p-value of t-test is greater than alpha, so we do not reject H0")
}
```
Since a p-value of `r t_test_pvalue` is greater than than $\alpha=0.05$, we cannot reject $H_0$ that $\mu_C \geq \mu_G$, and conclude that the expected value of bill lengths of Chinstrap penguins is greater than that of Gentoo penguins.

\pagebreak

# Exercise 2
### Q1
```{r}
species_na_bool_idx <- is.na(penguins_df$species)
body_mass_na_bool_idx <- is.na(penguins_df$body_mass_g)
combined_bool_idx <- !species_na_bool_idx & !body_mass_na_bool_idx

filtered_penguins_df <- penguins_df[combined_bool_idx, ]
rows_removed_num <- nrow(penguins_df) - nrow(filtered_penguins_df)
```
There are `r rows_removed_num` rows removed due to missing `species` or `body_mass` values.

### Q2
```{r}
species <- unique(filtered_penguins_df$species)
species_color_list <- c("red", "blue", "green")
combined_species_df <- data.frame()
for (species_i in species) {
  species_idx <- filtered_penguins_df$species == species_i
  species_df <- filtered_penguins_df[species_idx, ]
  species_df_slice <- species_df[c("species", "body_mass_g")]
  combined_species_df <- rbind(combined_species_df, species_df_slice)
}

ggplot(combined_species_df, aes(body_mass_g, fill = species)) +
  geom_density(alpha = 0.2) +
  xlim(2000, 7000) +
  ylim(0, 1.25e-3) +
  ggtitle("Density of Body Mass (g) by Species")
```
It appears that Chinstrap and Adelie pengins have a similar average weight, which is lower than that of Gentoo penguins. Adelie and Gentoo penguins also have wider variance in weight than Chinstrap penguins.

### Q3
```{r}
filtered_penguins_df$body_mass_kg <- filtered_penguins_df$body_mass_g / 1000
neg.logl <- function(theta, pi1, pi2, pi3, w1, w2, w3, x) {
  mu1 <- theta[1]
  mu2 <- theta[2]
  mu3 <- theta[3]

  sigma_1 <- exp(theta[4])
  sigma_2 <- exp(theta[5])
  sigma_3 <- exp(theta[6])
  # density of the mixture model:
  f.x1 <- pi1 * dnorm(x, mu1, sd = sigma_1)
  f.x2 <- pi2 * dnorm(x, mu2, sd = sigma_2)
  f.x3 <- pi3 * dnorm(x, mu3, sd = sigma_3)
  # negative log-likelihood:
  -sum(w1 * log(f.x1) + w2 * log(f.x2) + w3 * log(f.x3))
}
```

### Q4
```{r}
set.seed(3953629)

em_algo <- function(x, pi1hat_init, pi2hat_init, pi3hat_init, n.iter) {
  # Set n as length of provided x
  n <- length(x)

  # Set up pi_j for each j component
  pi1hat <- rep(NA, n.iter)
  pi2hat <- rep(NA, n.iter)
  pi3hat <- rep(NA, n.iter)

  # Set up p_hat_j for each j component
  p1hat <- matrix(NA, n.iter, n)
  p2hat <- matrix(NA, n.iter, n)
  p3hat <- matrix(NA, n.iter, n)

  # Define starting range for pi_j
  pi1hat_range <- pi1hat_init * 0.8
  pi2hat_range <- pi2hat_init * 0.8
  pi3hat_range <- pi3hat_init * 0.8

  # Store initial provided pi values
  pi1hat[1] <- pi1hat_init
  pi2hat[1] <- pi2hat_init
  pi3hat[1] <- pi3hat_init

  # Calculate t=1 p_hats
  p1hat[1, ] <- runif(n, pi1hat_init - pi1hat_range, pi1hat_init + pi1hat_range)
  p2hat[1, ] <- runif(n, pi2hat_init - pi2hat_range, pi2hat_init + pi2hat_range)
  p3hat[1, ] <- runif(n, pi3hat_init - pi3hat_range, pi3hat_init + pi3hat_range)

  # Set up 6 column theta matrix, and t=1 theta guesses
  thetahat <- matrix(NA, n.iter, 6)
  theta_init_guesses <- c(
    rep(mean(x), 3),
    rep(sd(x), 3)
  )

  # Find initial log likelihood minimized values of theta
  thetahat[1, ] <- optim(
    theta_init_guesses,
    function(theta) {
      neg.logl(
        theta, pi1hat[1], pi2hat[1], pi3hat[1],
        p1hat[1, ], p2hat[1, ], p3hat[1, ], x
      )
    }
  )$par

  # Iter through n.iter
  for (t in 2:n.iter) {
    # E step: update individual probability memberships
    phat_numer.temp <- cbind(
      pi1hat[t - 1] * dnorm(x, thetahat[t - 1, 1], exp(thetahat[t - 1, 4])),
      pi2hat[t - 1] * dnorm(x, thetahat[t - 1, 2], exp(thetahat[t - 1, 5])),
      pi3hat[t - 1] * dnorm(x, thetahat[t - 1, 3], exp(thetahat[t - 1, 6]))
    )
    # Reassign phat values to corresponding phat matrix
    p1hat[t, ] <- phat_numer.temp[, 1] / rowSums(phat_numer.temp)
    p2hat[t, ] <- phat_numer.temp[, 2] / rowSums(phat_numer.temp)
    p3hat[t, ] <- phat_numer.temp[, 3] / rowSums(phat_numer.temp)
    # M step: update parameter estimates
    # Update for corresponding pi_hat
    pi1hat[t] <- mean(p1hat[t, ])
    pi2hat[t] <- mean(p2hat[t, ])
    pi3hat[t] <- mean(p3hat[t, ])
    # Update optimized theta values
    thetahat[t, ] <- optim(thetahat[t - 1, ], function(theta) {
      neg.logl(
        theta, pi1hat[t], pi2hat[t], pi3hat[t],
        p1hat[t, ], p2hat[t, ], p3hat[t, ], x
      )
    })$par
  }
  # Return final theta values, pi values, and phat values
  # final_theta <- thetahat[n.iter, ]
  # final_pi <- c(pi1hat[n.iter], pi2hat[n.iter], pi3hat[n.iter])
  return(list(
    theta = thetahat,
    pi1 = pi1hat,
    pi2 = pi2hat,
    pi3 = pi3hat,
    p1 = p1hat,
    p2 = p2hat,
    p3 = p3hat
  ))
}

n <- nrow(filtered_penguins_df)
peng_species_prop <- filtered_penguins_df %>%
  group_by(species) %>%
  count()
peng_species_prop$prop <- peng_species_prop$n / n

pi1hat_1 <- 1 / 3
pi2hat_1 <- 1 / 3
pi3hat_1 <- 1 / 3

pi1hat_2 <- c(peng_species_prop$prop)[1]
pi2hat_2 <- c(peng_species_prop$prop)[2]
pi3hat_2 <- c(peng_species_prop$prop)[3]

pi1hat_3 <- runif(1, 0, 0.45)
pi2hat_3 <- runif(1, 0, 0.45)
pi3hat_3 <- 1 - pi1hat_3 - pi2hat_3

n.iter <- 500

run_1 <- em_algo(
  filtered_penguins_df$body_mass_kg,
  pi1hat_1,
  pi2hat_1,
  pi3hat_1,
  n.iter
)
run_2 <- em_algo(
  filtered_penguins_df$body_mass_kg,
  pi1hat_2,
  pi2hat_2,
  pi3hat_2,
  n.iter
)
run_3 <- em_algo(
  filtered_penguins_df$body_mass_kg,
  pi1hat_3,
  pi2hat_3,
  pi3hat_3,
  n.iter
)
```
The starting points for `run_1` assume even probability across all three penguin distributions.

The starting points for `run_2` follows the empirical frequency distribution in the data.

THe starting points for `run_3` are randomized, ensuring all three $\pi_j$ still add to 1.

### Q5
```{r}
actual_props <- c(peng_species_prop$prop)

run_1_props <- list(
  Adelie = sum(run_1$p1[n.iter, ] > 0.5) / n,
  Gentoo = sum(run_1$p2[n.iter, ] > 0.5) / n,
  Chinstrap = sum(run_1$p3[n.iter, ] > 0.5) / n
)
run_2_props <- list(
  Chinstrap = sum(run_2$p1[n.iter, ] > 0.5) / n,
  Adelie = sum(run_2$p2[n.iter, ] > 0.5) / n,
  Gentoo = sum(run_2$p3[n.iter, ] > 0.5) / n
)
run_3_props <- list(
  Adelie = sum(run_3$p1[n.iter, ] > 0.5) / n,
  Gentoo = sum(run_3$p2[n.iter, ] > 0.5) / n,
  Chinstrap = sum(run_3$p3[n.iter, ] > 0.5) / n
)

adelie_comp <- run_1_props$Adelie - c(
  run_1_props$Adelie,
  run_2_props$Adelie,
  run_3_props$Adelie
)
chins_comp <- run_1_props$Chinstrap - c(
  run_1_props$Chinstrap,
  run_2_props$Chinstrap,
  run_3_props$Chinstrap
)
gentoo_comp <- run_1_props$Gentoo - c(
  run_1_props$Gentoo,
  run_2_props$Gentoo,
  run_3_props$Gentoo
)

knitr::kable(rbind(adelie_comp, chins_comp, gentoo_comp),
  col.names = c("Run 1", "Run 2", "Run 3")
)
```
As such, Run 1 or Run 3 would be best given that they return the closest possible proportion relative to the intial observed values. Moving forward, Run 1 will be used.

### Q6
```{r fig.width = 7, fig.height = 3}
pi_labels <- c("pi1", "pi2", "pi3")
for (pi_lab in pi_labels) {
  plot(run_1[[pi_lab]],
    xlab = "iteration",
    ylab = "pi",
    pch = 4,
    col = rgb(0, 0, 1, 0.3),
    ylim = c(0, 1),
    main = sprintf("Convergence of %s", pi_lab)
  )
  abline(h = run_1[[pi_lab]][n.iter], 
     col = 'red'
  )
}
```

Given stable behaviour after about 200 iterations, the algorithm seems to have converged.

### Q7
```{r}
pi_ests <- c(run_1$pi1[n.iter], run_1$pi2[n.iter], run_1$pi3[n.iter])
mu_ests <- run_1$theta[n.iter, 1:3]
sigma_ests <- run_1$theta[n.iter, 4:6]

mle_summary_df <- data.frame(rbind(pi_ests, mu_ests, sigma_ests),
                             row.names = c("pi", "mu", "sigma"))

knitr::kable(mle_summary_df,
             col.names = c("j = 1", "j = 2", "j = 3"),
             caption = "MLE of pi, mu, and sigma from Run 1")

mean_species_weights <- filtered_penguins_df %>%
                          group_by(species) %>%
                          summarise(weight = mean(body_mass_kg))
knitr::kable(mean_species_weights,
             col.names = c("Species", "Average Weight"),
             caption = "Average Weight (kg) by Species")
```
The weights of the components are not similar, with the first group (Adelie), representing the majority of penguins. 
The estimates of average weight for Adelie are quite close to the actual, but the other two species (Chinstrap and Gentoo), differ from the calculated averages by larger values.

### Q8
```{r fig.width = 7, fig.height = 4}
w1_prob <- run_1$p1[n.iter, ]
w2_prob <- run_1$p2[n.iter, ]
w3_prob <- run_1$p3[n.iter, ]

# par()
hist(w1_prob, 
     breaks=20, 
     col=rgb(1, 0, 0, 0.4),
     main = "Probability Distribution of Penguins being in W_j")
hist(w2_prob, 
     breaks=20, 
     col=rgb(0, 1, 0, 0.4),
     add=T)
hist(w3_prob, 
     breaks=20, 
     col=rgb(0, 0, 1, 0.4), 
     add=T)
legend("top",
       c("W_1", "W_2", "W_3"),
       ncol = 3,
       fill = c(rgb(1, 0, 0, 0.4),
                rgb(0, 1, 0, 0.4),
                rgb(0, 0, 1, 0.4))
       )

```

### Q9
```{r}
predictedComponent <- ifelse(run_1$p1[n.iter, ] > 0.5, 1, 0) + 
  ifelse(run_1$p2[n.iter, ] > 0.5, 2, 0) +
  ifelse(run_1$p3[n.iter, ] > 0.5, 3, 0)
print(predictedComponent)
```

### Q10
```{r}
sex_pc_jfd <- table(filtered_penguins_df$sex, predictedComponent)
knitr::kable(sex_pc_jfd,
             caption = "Joint Frequency Distribution of Sex vs. Predicted W_j Component")

species_pc_jfd <- table(filtered_penguins_df$species, predictedComponent)
knitr::kable(species_pc_jfd,
             caption = "Joint Frequency Distribution of Species vs. Predicted W_j Component")
```

### Q11
```{r}
actual_sex_species <- table(filtered_penguins_df$sex, filtered_penguins_df$species)
knitr::kable(actual_sex_species,
             caption = "Joing Frequency Distribution of Sex vs. Species")
```
If a penguin is in PC1, it is most likely a female, as there is a `r sex_pc_jfd[1, 1]/sum(sex_pc_jfd[, 1])` probability of female from PC1.
If a penguin is in PC2, it is most likely a female, as there is a `r sex_pc_jfd[1, 2]/sum(sex_pc_jfd[, 2])` probability of female from PC2.
If a penguin is in PC3, it will be male, as there is a `r sex_pc_jfd[2, 3]/sum(sex_pc_jfd[, 3])` probability of male from PC3 - in short there are no females in PC3.

These guesses would be moderately accurate for PC1 and PC2, but far from reality from PC3 given the actual sex vs. species frequency distribution above.

### Q12
Based on the above JDF between PC and species, PC1 would align with Adelie, PC2 would align with Chinstrap, and PC3 would align with Gentoo 